Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2024-06-04 paper/ms.tex
finished a round of editing of paper/ms

\newcommand\ei[1]{\textcolor{red}{#1}}

@@

We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$, and we write writing $\texttt{process}\left(x_n\mid x_{n-1}; \theta\right)$ for its corresponding simulator.
We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, and $y_n^*$ for the actual values of the observations that were observed.
We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function \ei{NOT DEFINED YET. THIS WILL PUZZLE READERS UNLESS THEY ARE ALREADY HIGHLY FAMILIAR WITH THE TOPIC}.
For simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0$. \ei{ALGORITHM 1 HAS $f_{X_0}$. IN GENERAL, THIS DETAIL IS BEST INCLUDED. INITIAL VALUES CAN BE AN IMPORTANT PART OF A DYNAMIC SYSTEM.}

The above conditional densities are defined on a probability space $(\Omega, \Sigma, \prob).$ 
For the algorithmic interpretation of our theory, we identify the seed with an element of the sample space $\omega \in \Omega$.
\ei{THE SEED CORRESPONDS TO ALL THE MONTE CARLO QUANTITIES GENERATED, INCLUDING THINGS THAT HAVE NO DIRECT ROLE IN THE MODEL, SO MAYBE THIS EXPLANATION NEEDS ADDITIONAL CARE.}
The random seed determines the entire sequence of pseudo-random numbers as generated by a computer, just as the outcome $\omega\in\Omega$ generates the entire sequence of random variables.

@@

		\textbf{Initialize } particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, weights $w^{F,\theta}_{0,j}= 1$. \newline
		\textbf{For} $n=1,...,N$: \newline
            \hspace*{4mm} Accumulate discounted weights, $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$.\newline
            \hspace*{4mm} Simulate process model,
            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$. \newline
            \hspace*{4mm} Measurement density,
            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$. \newline
            \hspace*{4mm} Compute $L_n^{B,\theta,\alpha} ={\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$. \newline
            \hspace*{4mm} Conditional likelihood under $\phi$,
            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$.\newline
            \hspace*{4mm} Select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \ei{$\phi$ SHOULD BE $\theta$} \newline
            \hspace*{4mm} Resample corrected weights
            $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j}  { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}$.\newline
            \hspace*{4mm} Compute $ L_n^{A,\theta,\alpha} = L_n^\phi\cdot {\sum_{j=1}^J w^{F,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$.\newline
		\textbf{Return:} likelihood estimate $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, filtering distributions $\{(X_{n,j}^{F, \theta}, w^{F,\theta}_{n,j})\}_{n,j=1}^{N,J}.$
\end{algorithm}

\subsection{Algorithm Outline} 
We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but the resampling indices constructed from the first pass, with parameter $\phi$.
We call this ``off-policy'' measurement since a parameter value corresponds to a policy in the context of control theory.
\ei{I REWROTE THIS TO BE MORE EXPLICIT, BUT MAYBE THIS NEEDS MORE EDITING. IT MAKES ME THINK THAT ``RESAMPLING OFF-POLICY'' WOULD BE A BETTER NAME!}
If $\theta$ and $\phi$ coincide, one only needs one particle filter run at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.

We can then reweight the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$. That is, writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$, and $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$ for the resampling indices under $\phi$, we can estimate the conditional likelihood under $\theta$ as
\vspace*{-2.5mm}
\begin{equation}
     \label{eq:mop-conditional-likelihood}
     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}} \text{ or } L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
     \vspace*{-2.5mm}
\end{equation}
recursively weighting each particle by 
\vspace*{-2.5mm}
\begin{equation}
    \label{eq:weighting-scheme}
    w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha,  w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, w^{F,\theta}_{0,j}= 1.
    \vspace*{-2.5mm}
\end{equation}

@@

As mentioned earlier, the correction factors accumulate over time. If $\theta$ and $\phi$ coincide, one can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood at $\theta$. This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate. When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}. When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}.

@@

$\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories. This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.

@@

We defer the proof to the supplementary material. It relies on a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case. Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting that when $\theta=\phi$, that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.

This further illustrates how $\alpha$ dictates the "memory" of the gradient estimate. As \cite{scibior21} remark, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is "memoryless" beyond a single step. This is in contrast to the case when $\alpha=1,$ as that estimate of \cite{poyiadjis11} only considers the surviving particles at time $N$, fully tracking dependencies over time.

@@

We bypass the issue of differentiating through a Monte Carlo algorithm with discontinuous resampling by turning it into a problem of differentiating through a simulator and a series of measurement density ratios. Although we defer the theoretical analysis of the resulting algorithm, MOP-$\alpha$, to Section \ref{sec:thms}, we highlight a few key results below. MOP-$\alpha$ targets the posterior and likelihood (Theorem \ref{thm:mop-targeting}), yields the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases when differentiated (Theorem \ref{thm:mop-functional-forms}) where MOP-$1$ in particular is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).

\section{Practical Maximum-Likelihood Estimation}

@@

This still leaves us with the question of designing an effective procedure for likelihood maximization with MOP-$\alpha$. We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs a few iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate. This leverages IF2's quick empirical convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima. Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty that IF2 has with optimizing the last few units of log-likelihood. Combining these two methods in this way lets us enjoy the best of both worlds.
\kevin{source for this}

\subsection{Linear Convergence Rates}
\begin{thm}
    \label{thm:mop-convergence}
\end{thm}

%The warm start allows gradient methods to shine, as regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.

\begin{algorithm}[H]
	\caption{IFAD}
    \label{alg:ifad}
	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $t=0.$\newline
        Run IF2 until initial "convergence" under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
		\textbf{While} procedure not converged: \newline
		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_t).$ \newline
		\hspace*{4mm} Obtain $g(\theta_t) = \nabla_{\theta_t} (-\hat\loglik(\theta_t))$, $H(\theta_t)$ s.t. $\lambda_{\min}(H) \geq c$. \newline
		\hspace*{4mm} Update $\theta_{t+1} := \theta_t - \eta (H(\theta_t))^{-1} g(\theta_t)$, $t:=t+1.$ \newline
		\textbf{Return} $\hat{\theta} := \theta_t.$
\end{algorithm}

\section{Application to the Cholera Model Of \cite{king08}}

We use the Dhaka cholera model of \cite{king08} to benchmark (also used or modified in \cite{ionides15, wood16, wycoff24} for the same purpose) the performance of our method. This is a stochastic SIR compartmental model where the population at time $t$, $H_t$, is divided into the susceptible compartment $S_t$, infected compartment $I_t$, and three ($k=3$) recovered compartments $R^1_t, ..., R^k_t$ denoting varying degrees of cholera immunity. We write $M_t$ for the cholera deaths in each month. As in \cite{king08, ionides15}, the \textbf{transition dynamics} follow this series of SDEs:
\vspace*{-1mm}
\begin{align*}
    d S_t&=(k \epsilon R^k_t+\delta(S_t-H_t)-\lambda_t S_t) d t+d H_t-({\sigma S_t I_t}/{H_t}) d B_t, \\
    d I_t&=\left(\lambda_t S_t-(m+\delta+\gamma) I_t\right) d t+({\sigma S_t I_t}/{H_t}) d B_t, \\
    d R^1_t&=(\gamma I-(k \epsilon+\delta) R^1) d t,..., 
    d R^k_t=(k \epsilon R^{k-1}_t\hspace{-3mm}-(k \epsilon+\delta) R^k_t) d t,
    \vspace*{-2mm}
\end{align*}

with Brownian motion $B_t$, cholera death rate $m$, recovery rate $\gamma$, mean immunity duration $1/\epsilon$, standard deviation of the force of infection $\sigma$, and population death rate $\delta=0.02$. The \textbf{force of infection}, $\lambda_t$, is modeled by splines $(s_j)_{j=1}^6$
\vspace*{-2mm}
\begin{equation*}
    \lambda_t=\exp\hspace{-1mm}\left(\hspace{-.5mm}\beta_{\text{trend}}(t-t_0)+\hspace{-1mm}\sum_{j=1}^{6} \beta_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm}\frac{I_t}{P_t} + \exp\hspace{-1mm} \left(\sum_{j=1}^{6} \omega_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm},
    \vspace*{-2mm}
\end{equation*}
where the coefficients $(\beta_j)_{j=1}^6$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease. The \textbf{measurement model} for observed monthly cholera deaths is given by 
    $Y_n \sim \gN(M_n, \tau^2M_n^2)$,
where $M_n$ is the true number of cholera deaths in that month.

@@

We benchmarked IFAD against IF2 on a challenging global search problem. We re-implemented IF2 in order to do so, but chose to benchmark our results with the results of \cite{ionides15} (labeled "IF2 2015"). We note that our re-implementation ourperforms that of \cite{ionides15}, likely due to a better choice of parameters. For each method, we performed 100 searches, initialized with 100 initial starting parameter vectors drawn uniformly from the same wide bounding box used in \cite{ionides15}. We summarize our findings below.

@@

@@

\paragraph{IFAD Successfully Finds the MLE:} Previously, the MLE at a log-likelihood of $-3748.5$ was found by \cite{king16} with much effort, using many global and local IF2 searches, with the assistance of likelihood profiling. Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while \cite{king08} only achieve a best possible estimate of $-3793.4$. Despite being initialized for a global search, IFAD manages to find the MLE over the 100 searches, as seen in Table \ref{table:mle}. This suggests that the local search and refinement that was previously necessary for finding the MLE with IF2 is \textbf{baked into} the two-stage IFAD algorithm.

@@

We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood. While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in the supplementary material. That is, we show a strong law of large numbers for triangular arrays of particles with ``off-policy'' resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest, to weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.

@@

Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}, X_{1:N}}\left(y_{1:N}^* , x_{1:n,j}^{A, F,\theta}; \theta\right)$ in general. The bootstrap and general estimates have different functional forms, and are not the same a-priori. As such we directly show the consistency of the MOP-$1$ gradient estimate below.

@@

Re-implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python. This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.

>>>>

@@

We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$, and we write $\texttt{process}\left(x_n\mid x_{n-1}; \theta\right)$ for its corresponding simulator \ed{WE DON'T USE THIS... MAYBE CHANGE HERE, OR USE IT}.
We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, and $y_n^*$ for the actual values of the observations that were observed.
We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function \ed{NOT DEFINED YET. THIS WILL PUZZLE READERS UNLESS THEY ARE ALREADY HIGHLY FAMILIAR WITH THE TOPIC}.
For simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0$. \ed{ALGORITHM 1 HAS $f_{X_0}$. IN GENERAL, THIS DETAIL IS BEST INCLUDED. INITIAL VALUES CAN BE AN IMPORTANT PART OF A DYNAMIC SYSTEM.}

The above conditional densities are defined on a probability space $(\Omega, \Sigma, \prob).$ 
For the algorithmic interpretation of our theory, we identify the seed with an element of the sample space $\omega \in \Omega$.
\ed{THE SEED CORRESPONDS TO ALL THE MONTE CARLO QUANTITIES GENERATED, INCLUDING THINGS THAT HAVE NO DIRECT ROLE IN THE MODEL, SO MAYBE THIS EXPLANATION NEEDS ADDITIONAL CARE.}
The random seed determines the entire sequence of pseudo-random numbers as generated by a computer, just as the outcome $\omega\in\Omega$ generates the entire sequence of random variables.

@@

		\textbf{Initialize } particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, weights $w^{F,\theta}_{0,j}= 1$. \newline
		\textbf{For} $n=1,...,N$: \newline
            \hspace*{4mm} Accumulate discounted weights, $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$.\newline
            \hspace*{4mm} Simulate process model,
            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$. \newline
            \hspace*{4mm} Measurement density,
            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$. \newline
            \hspace*{4mm} Compute $L_n^{B,\theta,\alpha} ={\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$. \newline
            \hspace*{4mm} Conditional likelihood under $\phi$,
            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$.\newline
            \hspace*{4mm} Select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \ed{$\phi$ SHOULD BE $\theta$} \newline
            \hspace*{4mm} Resample corrected weights
            $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j}  { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}$.\newline
            \hspace*{4mm} Compute $ L_n^{A,\theta,\alpha} = L_n^\phi\cdot {\sum_{j=1}^J w^{F,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$.\newline
		\textbf{Return:} likelihood estimate $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, filtering distributions $\{(X_{n,j}^{F, \theta}, w^{F,\theta}_{n,j})\}_{n,j=1}^{N,J}.$
\end{algorithm}

\subsection{Algorithm Outline} 
We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but the resampling indices constructed from the first pass, with parameter $\phi$.
We call this ``off-policy'' measurement since a parameter value corresponds to a policy in the context of control theory.
\ed{I REWROTE THIS TO BE MORE EXPLICIT, BUT MAYBE THIS NEEDS MORE EDITING. IT MAKES ME THINK THAT ``RESAMPLING OFF-POLICY'' WOULD BE A BETTER NAME!}
The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
For the categorical distribution, we use systematic resamplig \cite{arulampalam02,king16} which usually has superior performance to multinomial resampling.
If $\theta$ and $\phi$ coincide, one only needs one particle filter run at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.

We then reweight the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$. That is, writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$,  we can estimate the conditional likelihood under $\theta$ as
\vspace*{-2.5mm}
\begin{equation}
     \label{eq:mop-conditional-likelihood}
     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}} \text{ or } L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
     \vspace*{-2.5mm}
\end{equation}
recursively weighting each particle by 
\vspace*{-2.5mm}
\begin{equation}
    \label{eq:weighting-scheme}
    w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha,  w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, w^{F,\theta}_{0,j}= 1.
    \vspace*{-2.5mm}
\end{equation}

@@

The correction factors, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulate as $n$ increases, leading to numerical instability.
One can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood if $\theta=\phi$.
This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, and Theorem~\ref{thm:mop-functional-forms} shows that it recovers the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}.

@@

$\alpha$ therefore controls an exponentially-weighted moving average, determining how much memory the resulting gradient estimate has over ancestral trajectories.
This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.

@@

We defer the proof to the supplementary material. It relies on a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case. Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.

This further illustrates how $\alpha$ dictates the memory of the gradient estimate. As \cite{scibior21} remark, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is ``memoryless'' beyond a single step. This is in contrast to the case when $\alpha=1,$ as that estimate, studied by \cite{poyiadjis11}, only considers the surviving particles at time $N$ and so fully tracks dependencies over time.

@@

The construction of MOP-$\alpha$ bypasses the issue of differentiating through a Monte Carlo algorithm with discontinuous resampling by turning it into a problem of differentiating through a simulator and a series of measurement density ratios.
We defer the theoretical analysis of MOP-$\alpha$ to Section \ref{sec:thms}, but first we highlight a few key results.
MOP-$\alpha$ estimates the likelihood and conditional distributions of latent variables, as one expects of a particle filter (Theorem \ref{thm:mop-targeting}).
When differentiated, we obtain the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases (Theorem \ref{thm:mop-functional-forms}).
In particular, MOP-$1$ is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).

\section{Practical Maximum Likelihood Estimation}

@@

We now consider the question of designing an effective procedure for likelihood maximization with MOP-$\alpha$.
We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs some preliminary iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate.
This leverages IF2's empirically stable and rapid convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima.
Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty experienced with IF2 when optimizing the last few units of log-likelihood.
Combining these two methods lets us enjoy the best properties of both algorithms, which we show leads to strong empirical performance.

\subsection{Linear Convergence Rates}
\begin{thm}
  \label{thm:mop-convergence}
  \ed{CURRENTLY EMPTY}
\end{thm}

\ed{IN IFAD, $t$ IS USED AS AN INDEX FOR ITERATIONS. PREVIOUSLY, WE HAVE USED $m$ FOR THIS SINCE $n$ AND $t$ ARE USED ELSEWHERE.}

\ed{IFAD HERE IS A 2ND ORDER ALGORITHM. MAYBE FOCUS ON THE PRIMARY EXAMPLE, WHICH IS FIRST ORDER. ALSO, THE STOPPING RULE SHOULD DESCRIBE WHAT WAS DONE IN THE EXEMPLES}

\begin{algorithm}[H]
	\caption{IFAD}
    \label{alg:ifad}
	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $t=0.$\newline
        \textbf{Preliminary iterated filtering:} Run IF2 under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
		\textbf{While} procedure not converged: \newline
		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_t).$ \newline
		\hspace*{4mm} Obtain $g(\theta_t) = \nabla_{\theta_t} (-\hat\loglik(\theta_t))$, $H(\theta_t)$ s.t. $\lambda_{\min}(H) \geq c$. \newline
		\hspace*{4mm} Update $\theta_{t+1} := \theta_t - \eta (H(\theta_t))^{-1} g(\theta_t)$, $t:=t+1.$ \newline
		\textbf{Return} $\hat{\theta} := \theta_t.$
\end{algorithm}

\section{Application to a Cholera Transmission Model}

The cholera transmission model that \cite{king08} developed for Dhaka, Bangladesh, has been used to benchmark the performance of various POMP inference methods \cite{ionides15, wood16, wycoff24}, and we employ it here for the same purpose.
This model categorizes individuals in a population as susceptible, $S(t)$, infected, $I(t)$, and recovered, $R(t)$ and so is called an SIR compartmental model.
In this case, the compartment $R(t)$ is further subdivided into 
%. where the population at time $t$, $H_t$, is divided into the susceptible compartment $S_t$, infected compartment $I_t$, and
three recovered compartments $R^1(t)$, $R^2(t)$, $R^3(t)$ denoting varying degrees of cholera immunity.
We write $H(t)$ for the total population, and $M_t$ \ed{$M_n$ ??} for the cholera deaths in each month.
As in \cite{king08, ionides15}, the transition dynamics follow a series of stochastic differential equations:
\ed{CHANGE TO $S(t)$ etc FOR CONSISTENCY.}
\vspace*{-1mm}
\begin{align*}
    d S_t&=(k \epsilon R^k_t+\delta(S_t-H_t)-\lambda_t S_t) d t+d H_t-({\sigma S_t I_t}/{H_t}) d B_t, \\
    d I_t&=\left(\lambda_t S_t-(m+\delta+\gamma) I_t\right) d t+({\sigma S_t I_t}/{H_t}) d B_t, \\
    d R^1_t&=(\gamma I-(k \epsilon+\delta) R^1) d t,..., 
    d R^k_t=(k \epsilon R^{k-1}_t\hspace{-3mm}-(k \epsilon+\delta) R^k_t) d t,
    \vspace*{-2mm}
\end{align*}
with Brownian motion $B_t$, cholera death rate $m$, recovery rate $\gamma$, mean immunity duration $1/\epsilon$, standard deviation of the force of infection $\sigma$, and population death rate $\delta=0.02$. The force of infection, $\lambda_t$, is modeled by splines $(s_j)_{j=1}^6$
\vspace*{-2mm}
\begin{equation*}
    \lambda_t=\exp\hspace{-1mm}\left(\hspace{-.5mm}\beta_{\text{trend}}(t-t_0)+\hspace{-1mm}\sum_{j=1}^{6} \beta_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm}\frac{I_t}{P_t} + \exp\hspace{-1mm} \left(\sum_{j=1}^{6} \omega_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm},
    \vspace*{-2mm}
\end{equation*}
where the coefficients $(\beta_j)_{j=1}^6$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
The measurement model for observed monthly cholera deaths is given by 
    $Y_n \sim \gN(M_n, \tau^2M_n^2)$,
where $M_n=\gamma\int_{t_{n-1}}^{t_n}I(s)\, ds$ is the modeled number of cholera deaths in that month.

@@

We tested IFAD against IF2 on a global search problem for the Dhaka cholera model.
We re-implemented IF2 in order to do so, but we compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
Our re-implementation ourperforms that of \cite{ionides15}, likely due to a better choice of algorithmic parameters.
For each method, we performed 100 searches, initialized with 100 initial starting parameter vectors drawn uniformly from the same wide bounding box used in \cite{ionides15}. We summarize our findings below.

@@

\ed{PERHAPS THE 2ND D.P. OF SIGNIFICANCE CAN BE AVOIDED? ALSO, IF WE'RE GOING TO ADVISE MOP-ALPHA FOR $0<\alpha<1$, THAT COULD BE SHOWN HERE.?}

@@

\paragraph{IFAD Successfully Finds the MLE:} Previously, the MLE at a log-likelihood of $-3748.5$ was found by \cite{king16} with much effort, using many global and local IF2 searches, with the assistance of likelihood profiling. Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while the best log-likelihood found by \cite{king08} was only $-3793.4$. Despite being initialized for a global search, IFAD manages to find the MLE over the 100 searches, as seen in Table \ref{table:mle}. This suggests that the sequence of local searches and refinement that was previously necessary for finding the MLE with IF2 is not necessary with the IFAD algorithm.

@@

We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood when $\alpha=1$ or $\theta=\phi$. While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in the supplementary material. That is, we show a strong law of large numbers for triangular arrays of particles with ``off-policy'' resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest, to weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.

@@

Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}, X_{1:N}}\left(y_{1:N}^* , x_{1:n,j}^{A, F,\theta}; \theta\right)$ in general. The bootstrap and general estimates have different functional forms, and are not the same a-priori \ed{WHAT DOES THIS MEAN? THAT IT IS NOT IMMEDIATLY CLEAR THEY ARE THE SAME?}. As such we directly show the consistency of the MOP-$1$ gradient estimate below.

@@

Implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, with a simulator written in Python. This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter in the \texttt{pomp} package of \cite{king16}, for which the simulator is written in C and called from R.
